Multi-level modeling with Openflexo/\FML - A contribution to the \mpc

### 2021-05-31 10-conclusions.tex
remove all capital letters from files, as specified by the style

As a future work we envision to explore a number of alternative solutions: first, we intend to propose another approach based on a more agnostic level solution such as Clabjects. Instead of two levels defined by Type and Instance concepts, we could have a single concept mixing a type part and an instance part. This approach would probably be easier if specifications evolve. Secondly, another possible approach is the use of the free modeling tool proposed by Openflexo. The solution would be developed from examples (Acme and XSure processes for instance) from which models would be identified. Finally, the realisation of this challenge highlighted the interest of integrating specific behaviors for multi-level concepts management in Openflexo infrastructure, to avoid tailoring them, as in this exercise. We intend to do so.

>>>>

As a future work we envision to explore a number of alternative solutions: first, we intend to propose another solution based on a more level-agnostic approach such as Clabjects. Instead of two levels defined by Type and Instance concepts, we could have a single concept mixing a type part and an instance part. This approach would probably ease adaptation if the problem specifications evolve. Secondly, another possible approach is the use of the free modeling tool proposed by Openflexo. The solution would be developed from examples (Acme and XSure processes for instance) from which models would be identified. Finally, the realisation of this challenge highlighted the interest of integrating specific behaviors for the management of multi-level concepts in the Openflexo infrastructure (as opposed to tailoring them as we did in this exercise). We intend to do so.

### 2021-05-31 4-technology.tex
remove all capital letters from files, as specified by the style

To meet the \mpc, we have decided to use the general
approach of model federation~\cite{Golra2016-federation}. \emph{Model
  federation} is a way to assemble models using some kind of
low-coupling links. It has been studied to answer the \OMG' RFP
Semantic Modeling for Information Federation (SIMF)~\cite{simf}. In
contrast to approaches that compose metamodels into a single large
metamodel grouping all needed entities, model federation build links
among models and metamodels (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/metamodel conformity -- that is presented
in~\cite{models2016-freemodel}. Another very interesting feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.

We decided to use this approach since it offers the possibility to
link and to navigate among levels. Before we describe the overall
architecture in next section, here are some key concepts implemented
in the Openflexo~\cite{openflexo_link} framework.

@@

This framework relies on the architecture of Figure~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Flexo Modeling
Language (\FML). Each federated model can be viewed as an autonomous
component that may evolve with its own tooling. The virtual models
serve as control components.
In this paper, we mainly use the conceptual models and more precisely \FML its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

@@

\FML simplified metamodel is provided in Figure~\ref{fig:mm}. It is designed to reify models as virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand \FML\footnote{Some aspects of \FML do not exist in object oriented approach.}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
These roles have types defining the kind of value the role will point
at runtime.
Whenever a type external to the federation space is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity in charge of giving
access to external elements using
a \emph{technology adapter}\footnote{It is a reusable library that defines
connections between the \FML execution engine and a particular
technological space.}.

\FML is designed to define not only the structure of virtual models but
also to define the collection of actions an engineer can perform on
them. These actions are called \emph{behaviors} and can either be called (like usual methods) or triggered by events. The reactive behaviors are mainly useful when models have to be synchronised. We did not
exploit this possibility in the challenge.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers a \FML
execution engine with an interactive virtual model design environment.
It has been used in several case studies including model mapping,
multi-paradigm process modeling and enterprise architecting. It has
also been used to build a tool, the freemodeling editor that has been
used in industrial projects~\cite{models2016-freemodel}. As of today, this tool
offers some mature technology adapters (\emph{docx} and \emph{excel}
for documents, \EMF and OWL for modeling languages, JDBC for databases,
REST and XMLRPC for external services and one for diagramming tools)
and other more rudimentary (pdf, HTTP, XML, etc.).

Finally, tools have their own model. We have taken advantage of
Openflexo features to build, in parallel with the models and virtual
models required by the challenge, a drawing tool that makes our
solution (partially) executable.

>>>>

To meet the \mpc, we have decided to use the general
approach of model federation~\parencite{Golra2016-federation}. \emph{Model
  federation} is a way to assemble models using some kind of
low-coupling links. It has been studied to answer the \OMG' RFP
Semantic Modeling for Information Federation~\parencite{simf}. In
contrast to approaches that compose metamodels into a single large
metamodel grouping all needed entities, model federation build links
among models and metamodels (even through levels) to make ``things''
work together. As an illustration of this feature, we developed a
free-modeling editor -- freeing oneself from the bonds of
model/metamodel conformity -- that is presented
in~\parencite{models2016-freemodel}. Another notable feature of
this approach is the strong decoupling among tools that remain usable
after federations are made.

We decided to use this approach since it offers the possibility to
link and to navigate among levels. Before we describe the overall
architecture in next section, here are some key concepts implemented
in the Openflexo~\parencite{openflexo_link} framework.

@@

This framework relies on the architecture of Figure~\ref{fig:mf}. A federation
gathers a set of conceptual models, named \emph{virtual models} and a
set of \emph{federated models}. Each federated model pertains to a
\emph{technological space} and uses the language of its specific
paradigm while a virtual model is built using the Flexo Modeling
Language (\FML). Each federated model is an autonomous
element that may evolve with its own tooling. The virtual models
serve as control elements binding the federated models together.
In this paper, we mainly use conceptual models and more precisely \FML its domain specific modeling language. The only use of the federation aspect is on the tooling made for the process language.

@@

\FML simplified metamodel is provided in Figure~\ref{fig:mm}. It is designed to define models as virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units forming architectures while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand \FML\footnote{Some aspects of \FML do not exist in object oriented approach.}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
These roles have types defining the kind of value the role will point
at runtime.
Whenever a type external to the federation space is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity in charge of giving
access to external elements using
a \emph{technology adapter}\footnote{It is a reusable library that defines
connections between the \FML execution engine and a particular
technological space.}.

\FML is designed to define not only the structure of virtual models but
also to define the collection of actions an engineer can perform on
them. These actions are called \emph{behaviors} and can either be called (like usual methods) or triggered by events. The reactive behaviors are mainly useful when a federated model evolution needs to trigger a computation. We do not
exploit this possibility in the challenge.

@@

The tool support for model federation framework,
Openflexo\footnote{\url{https://github.com/openflexo-team}}, is
developed as an open source initiative. This tool offers a \FML
execution engine with an interactive virtual model design environment.

Finally, tools have their own model. We have taken advantage of
Openflexo features to build, in parallel with the models  required by the challenge, a drawing tool that makes our
solution (partially) executable.

### 2021-05-31 7-requirementssatisfaction.tex
remove all capital letters from files, as specified by the style

\begin{itemize}
    \item \textbf{Conceptualization} (\textit{Conc.}): a concept carries the semantics exposed by the requirement (for example for \textbf{P1}: \textit{process type} is conceptualized in a \textit{FlexoConcept} \textit{ProcessType}).
    \item \textbf{Specialization} (\textit{Spec.}): a concept is specialized in the sense of object-oriented modeling (for example for \textbf{P2}: \textit{Gateway} defines an abstract behaviour \texttt{execute()} and \textit{Sequencing}, \textit{AndSplit}, \textit{AndJoin}, \textit{OrSplit} and \textit{OrJoin} are defined as inherited concepts implementing specific business logic).
    \item \textbf{Composition} (\textit{Comp.}): some concepts are in relation, or specific attributes were added to some concepts (for example for \textbf{P3}: a \textit{process type} has one \textit{initial task type}).
    \item \textbf{Linguistic instanciation} (\textit{L.Inst.}): requirement is satisfied by the instantiation of one or more concepts (for example for \textbf{S1}: \textit{requirement analysis} is defined as an instance of \textit{SETasktype}).
    \item \textbf{Ontologic instanciation} (\textit{O.Inst.}): fulfillment of requirement is obtained by the relation between an instance and its type definition instance (for example for \textbf{S3} : \textit{developer} has the dual concept nature in \textit{Acme metamodel} and instance nature in \textit{Acme model}).
    \item \textbf{Behavioral modeling} (\textit{B.Mod.}): requirement is satisfied by the definition of one more object-oriented behavioural features, which may combined with constraints checking and associated tooling (for example \textbf{P17}).
\end{itemize}

@@

%CT : Conceptualization
%SP : Specialization
%CP : Composition
%LI : Linguistic instantiation
%OI : Ontologic instantiation
%BM : Behavioural modeling

>>>>

\begin{itemize}
    \item \textbf{Conceptualization} (\textit{Conc.}): a concept carries the semantics exposed by the requirement (for example for \textbf{P1}: \textit{process type} is conceptualized in a \textit{FlexoConcept} \textit{ProcessType}).
    \item \textbf{Specialization} (\textit{Spec.}): a concept is specialized in the sense of object-oriented modeling (for example for \textbf{P2}: \textit{Gateway} defines an abstract behavior \texttt{execute()} and \textit{Sequencing}, \textit{AndSplit}, \textit{AndJoin}, \textit{OrSplit} and \textit{OrJoin} are defined as inherited concepts implementing specific business logic).
    \item \textbf{Composition} (\textit{Comp.}): some concepts are in relation, or specific attributes were added to some concepts (for example for \textbf{P3}: a \textit{process type} has one \textit{initial task type}).
    \item \textbf{Linguistic instanciation} (\textit{L.Inst.}): requirement is satisfied by the instantiation of one or more concepts (for example for \textbf{S1}: \textit{requirement analysis} is defined as an instance of \textit{SETasktype}).
    \item \textbf{Ontologic instanciation} (\textit{O.Inst.}): fulfillment of requirement is obtained by the relation between an instance and its type definition instance (for example for \textbf{S3} : \textit{developer} has the dual concept nature in \textit{Acme metamodel} and instance nature in \textit{Acme model}).
    \item \textbf{Behavioral modeling} (\textit{B.Mod.}): requirement is satisfied by the definition of one more object-oriented behavioral features, which may combined with constraints checking and associated tooling (for example \textbf{P17}).
\end{itemize}

@@

%CT : Conceptualization
%SP : Specialization
%CP : Composition
%LI : Linguistic instantiation
%OI : Ontologic instantiation
%BM : behavioral modeling

### 2021-05-31 8-discussion.tex
remove all capital letters from files, as specified by the style

  \noteJC{pas clair pour moi ce qui est entendu par « basic modeling constructs », ce qui est dans la partie 2 ?}

>>>>

%  \noteJC{pas clair pour moi ce qui est entendu par « basic modeling constructs », ce qui est dans la partie 2 ?}

  Our solution uses the basic modeling constructs depicted by the \FML core
  metamodel (Figure~\ref{fig:mm}) and described in
  section~\ref{sec:technology}. Models are \emph{virtual models}. They are
  composed of \emph{concepts}, being themselves concepts. Concepts may have
  \emph{roles} and \emph{behaviors} (actions one can perform).

  In order to provide the graphical tools, our solution also uses \emph{model
  slots} to connect some concept instances to external elements (in our case:
  instance of diagram from our diagramming TA). The use of slots has been
  described in section~\ref{sec:model} and is illustrated by bold circles in
  the Figure~\ref{fig:ToolingArchitecture}).

### 2021-05-31 9-relatedworks.tex
remove all capital letters from files, as specified by the style

%Salvador commence
A plethora of multi-level modeling approaches and tools with different foundations have appeared in the recent years (see \cite{multiTools})\todo{not sure how to cite a wiki page}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

@@

\cite{multiecore2019} use MultiEcore \citep{multecore2016} to provide a solution to the challenge. MultEcore uses an extesion of the two-level cascading technique and potency. They do not need the synthetic typing relations we introduce. However they need to do frequent model transformations to transform instance models into instantiable models. With respect to them, we use less models as we do not include a generic software engineering process model nor a generic assurance process model. Finally, they use multi-level couple model transformations for the definition and enforcement of constraints.

More similar to us, \cite{deeptelos2019} use DeepTelos \citep{deeptelos2016}, an extension of the Telos language \citep{telos1990} in order to solve the process challenge. As with Openflexo and \FML, they do not use explicit (numbered) levels nor potency. However, DeepTelos is based on the use of the PowerType \citep{atkinson2001essence} pattern.

>>>>

%Salvador commence
A plethora of multi-level modeling approaches and tools with different foundations have appeared in the recent years~\footnote{\url{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/MultiTools}}. Comparing them lies out of the scope of this paper. In this sense, we limit the following discussion to the presentation and comparison of previous solutions to the MULTI Process Challenge.

@@

\cite{multiecore2019} use MultiEcore \citep{multecore2016} to provide a solution to the challenge. MultEcore uses an extension of the two-level cascading technique and potency. They do not need the synthetic typing relations we introduce. However they need to do frequent model transformations to transform instance models into instantiable models. With respect to the proposed solution, we use less models as we do not include a generic software engineering process model nor a generic assurance process model. %Finally, they use multi-level couple model transformations for the definition and enforcement of constraints.

More similar to us, \cite{deeptelos2019} use DeepTelos \citep{deeptelos2016}, an extension of the Telos language \citep{telos1990} in order to solve the process challenge. As with Openflexo and \FML, they do not use explicit (numbered) levels nor potency. However, unlike Openflexo, DeepTelos integrates a multi-level modeling specific construct similar to the PowerType \citep{atkinson2001essence} pattern they call \emph{most general instances}.
